Exploring Factors Influencing Average Playtime of Video Games: A Regression Analysis

### 2024-12-04 paper/paper.Rmd
Rendered paper into pdf.

---
title: "Exploring Factors Influencing Average Playtime of Video Games: A Regression Analysis"
subtitle: ""
author: 
  - name: "Ziqi Zhu"
  - name: "Raghav Sinha"
  - name: "Christophe McWatt"
thanks: "Code and data are available at: [https://github.com/zzq20010617/sta302Paper](https://github.com/zzq20010617/sta302Paper)."
date: today
date-format: long
abstract: ""
number-sections: true
bibliography: references.bib
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
---

@@

# Contributions
We found the dataset together, and seperate the works as follow:
Ziqi: Introduction and data description
Raghav: Ethics discussion and some plot in preliminary result
Christophe: Most of Preliminary results section

# Introduction
The gaming industry has experienced exponential growth in recent years, with millions of players engaging in various games on platforms like Steam. Understanding what influences average playtime in games is crucial for game developers, publishers, and marketers. Average playtime is a key indicator of player engagement and game success, and investigating factors that impact it can provide valuable insights into game design, pricing strategies, and marketing efforts. So our research aims to explore the factors that influence the average playtime of games available on Steam using a dataset we get from Kaggle. Specifically, we will investigate how variables such as price, user ratings, and other game characteristics contribute to player engagement, as measured by playtime.
From some previous papers, we learned that both positive and negative reviews have a significant correlation with playtime, but the effect is more pronounced for positive reviews [@paper1]. Another paper found out that game pricing significantly affects player engagement, with lower-priced games generally seeing higher playtime [@paper2]. The third paper did an empirical study that also examined the influence of game reviews on playtime, focusing on how the number and sentiment of reviews can affect player engagement. It concludes that games with more positive reviews see increased playtime, especially if those reviews highlight the game’s quality [@paper3].
We found this problem suitable for linear regression because it allows us to model the relationship between average playtime (the dependent variable) and various predictors (price, ratings, developer, etc.). This approach helps quantify how each factor influences playtime, enabling a clear understanding of the impact of different game attributes. By fitting a linear regression model, we can determine which factors have statistically significant effects on playtime, providing actionable insights for game developers and marketers.

@@

<!-- ## Fit multiple linear regression model -->
```{r}
#| echo: false
#| warning: false
#| message: false

# Load first model fit with lm(average_playtime ~ price + positive_ratings + 
# negative_ratings + achievements + owners, data=analysis_data)
model_1 <- readRDS(file = here::here("models/model_1.rds"))
```
# Results

At the starting point, our baseline model uses `average_playtime` as the response variable, with `price`, `positive_ratings`, `negative_ratings`, `achievements`, and `owners` as predictors. The four key linear regression assumptions considered are linearity, uncorrelated errors, constant error variance, and normality. Additionally, since this is a multiple linear regression model, we also examine the conditional mean response and predictor assumptions.
<!-- ## Create Response vs Fitted Scatterplot to check patterns in residuals before further analysis. -->
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-resp-fit
#| fig-cap: "Response vs. Fitted plot to assess model fit."

y_hat <- fitted(model_1)
plot(x = y_hat, y = data$average_playtime, main = "Response vs. Fitted", xlab = "Fitted", ylab = "Average Playtime")
abline(a=0, b=1, lty=2)
```
To check the conditional mean response assumption, we analyze the Response vs. Fitted scatterplot @fig-resp-fit. The data points are tightly clustered in the bottom-left corner, which satisfies this assumption. The fitted line generally follows the line $y=\hat{y}$, but there are notable outliers along the Y-axis where the observed `average_playtime` deviates significantly from the predicted values. These extreme outliers suggest that the response variable may be highly skewed or contain problematic data points.

@@

<!-- ## Create Pairwise Scatterplots for Predictors to look for multicollinearity -->
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-pairwise
#| fig-cap: "Pairwise plot between numeric predictors"

@@

<!-- ## Create Residuals vs Fitted Scatterplot to check Linearity, Uncorrelated Errors, and Constant Variance -->
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-resfit
#| fig-cap: "Residual vs fitted plot for model lm(average_playtime ~ price + positive_ratings + 
#|  negative_ratings + achievements + owners, data=data)"

@@

<!-- ## Create Normal QQ Plot to check Normality assumption -->
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-qq
#| fig-cap: "Residual vs fitted plot for model"

qqnorm(y_value)
qqline(y_value)
```
<!-- ## Box-Cox Transformation -->
```{r}
#| echo: false
#| warning: false
#| message: false

@@

```
<!-- ## Fitting transformed model -->
```{r}
#| echo: false
#| warning: false
#| message: false

@@

<!-- ## Checking for Multicollinearity -->
```{r}
#| include: false
#| echo: false
#| warning: false
#| message: false

vif(model_transformed)
```
To address this issue, `bc_negative_ratings` was removed from the model, leaving the final predictors as `bc_price` (log-transformed of `price`), `bc_positive_ratings`(log-transformed of `positive_ratings`),  `bc_achievements`(log-transformed of `achievements`) and `owners`. The refined model was re-evaluated with summary statistics and residual plots. The summary of coefficients as shown in @tbl-summary will be discussed in Section @sec-conclusion. The residual diagnostics @fig-reduced show significant improvement, with a more random distribution around zero in the Residuals vs. Fitted plot and a QQ plot with better alignment to the reference line.

<!-- ## Generatingn Residual Plots -->
```{r, fig.width=8, fig.height=6}
#| echo: false
#| warning: false
#| message: false
#| label: fig-reduced
#| fig-cap: "Diagnostic plots for lm(bc_average_playtime ~ bc_price + bc_positive_ratings + bc_achievements + owners, data=data)"

@@

# plot(x = x_value, y = y_value, main = "Residual vs. Fitted", xlab = "Fitted", ylab = "Residuals")
```
To further simplify the model, automated selection procedures using AIC in both forward and backward directions were conducted. The procedure resulted in the same final model as described above, confirming its robustness.

<!-- ## AIC Model Selection -->
```{r}
#| include: false
#| echo: false
#| warning: false
#| message: false

step_model <- stepAIC(reduced, direction = "both")
summary(step_model)
```
<!-- ## ANOVA Comparison between full model and reduced model -->
```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

@@

In conclusion, the model demonstrates that the average playtime of a game is positively influenced by several factors, including the game's price, the number of achievements, the number of positive reviews, and the estimated number of owners. Notably, the number of owners has the most substantial effect, as shown in the model summary @tbl-summary. For instance, games with an estimated ownership level of 100,000–200,000 have a coefficient of approximately 0.81; as ownership grows by a factor of 10, the effect increases to ~1.70, and for another 10-fold increase, the effect grows to ~3.04. This indicates that games with a larger player base tend to have higher average playtime, likely due to increased community engagement and multiplayer opportunities.

@@

The limitations in this study is ...

# Ethics discussion
<!-- The Steam Games dataset [@davis2023steam] used in this research is a collected dataset, not simulated, comprising information on over 97,000 games published on the Steam platform. The metadata is comprehensively filled out, including details such as game titles, release dates, genres, and user ratings, which enhances the dataset's usability and reliability. The source of the data is clearly described, originating from the Steam store pages and Steam API, ensuring transparency and traceability. Additionally, the dataset is hosted on Kaggle, a reputable platform for data science and machine learning, and has been vetted and used in existing literature (discussed in the introduction). This widespread use and accessibility suggest that the dataset is well-regarded and trusted by third parties, further validating its credibility through the framework outlined in the ethics module. There are no ethical concerns to report. -->

@@

# Appendix {-}

# Model details {#sec-model-detail}
## Residuals vs each Quantitative Predictor scatterplot to check Linearity, Uncorrelated Errors, and Constant Variance
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-respre
#| fig-cap: "Residuals vs each Quantitative Predictor"

@@

## BoxPlot to check assumptions for Categorical Predictors
See figure @fig-boc-cat
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boc-cat
#| fig-cap: "Boxplot residual vs. Owner levels suggest no linear assumption violation regarding the categorical predictors"

@@

boxplot(y_value ~ data$owners , main="Residual vs Owners",
xlab="Owners", ylab="Residuals")
```
## Final model results
<!-- ## Fitting Reduced Model -->
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary

@@

Checking for problematic points-leverage cutoff: ~0.006, cook's distance cutoff: ~0.961, dffits cutoff: ~0.105, dfbeta cutoff: ~0.025. Number of influential points by each case-leverage points: 216, regression outliers: 7, Cook's distance: 0, DFFITS: 211, by Beta from 0 to 7: 431, 476, 377, 394, 464, 321, 49, 0.
```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false

>>>>

---
title: "Exploring Factors Influencing Average Playtime of Video Games: A Regression Analysis"
subtitle: ""
author: 
  - name: "Ziqi Zhu"
  - name: "Raghav Sinha"
  - name: "Christophe McWatt"
thanks: "Code and data are available at: [https://github.com/zzq20010617/sta302Paper](https://github.com/zzq20010617/sta302Paper)."
date: today
date-format: long
abstract: ""
number-sections: true
bibliography: references.bib
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
editor: 
  markdown: 
    wrap: 72
---

@@

# Contributions

We found the dataset together, and seperate the works as follow: Ziqi:
Introduction and data description Raghav: Ethics discussion and some
plot in preliminary result Christophe: Most of Preliminary results
section

# Introduction

The gaming industry has experienced exponential growth in recent years,
with millions of players engaging in various games on platforms like
Steam. Understanding what influences average playtime in games is
crucial for game developers, publishers, and marketers. Average playtime
is a key indicator of player engagement and game success, and
investigating factors that impact it can provide valuable insights into
game design, pricing strategies, and marketing efforts. So our research
aims to explore the factors that influence the average playtime of games
available on Steam using a dataset we get from Kaggle. Specifically, we
will investigate how variables such as price, user ratings, and other
game characteristics contribute to player engagement, as measured by
playtime. From some previous papers, we learned that both positive and
negative reviews have a significant correlation with playtime, but the
effect is more pronounced for positive reviews [@paper1]. Another paper
found out that game pricing significantly affects player engagement,
with lower-priced games generally seeing higher playtime [@paper2]. The
third paper did an empirical study that also examined the influence of
game reviews on playtime, focusing on how the number and sentiment of
reviews can affect player engagement. It concludes that games with more
positive reviews see increased playtime, especially if those reviews
highlight the game’s quality [@paper3]. We found this problem suitable
for linear regression because it allows us to model the relationship
between average playtime (the dependent variable) and various predictors
(price, ratings, developer, etc.). This approach helps quantify how each
factor influences playtime, enabling a clear understanding of the impact
of different game attributes. By fitting a linear regression model, we
can determine which factors have statistically significant effects on
playtime, providing actionable insights for game developers and
marketers.

@@

<!-- ## Fit multiple linear regression model -->

```{r}
#| echo: false
#| warning: false
#| message: false

# Load first model fit with lm(average_playtime ~ price + positive_ratings + 
# negative_ratings + achievements + owners, data=analysis_data)
model_1 <- readRDS(file = here::here("models/model_1.rds"))
```

# Results

At the starting point, our baseline model uses `average_playtime` as the
response variable, with `price`, `positive_ratings`, `negative_ratings`,
`achievements`, and `owners` as predictors. The four key linear
regression assumptions considered are linearity, uncorrelated errors,
constant error variance, and normality. Additionally, since this is a
multiple linear regression model, we also examine the conditional mean
response and predictor assumptions.
<!-- ## Create Response vs Fitted Scatterplot to check patterns in residuals before further analysis. -->

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-resp-fit
#| fig-cap: "Response vs. Fitted plot to assess model fit."

y_hat <- fitted(model_1)
plot(x = y_hat, y = data$average_playtime, main = "Response vs. Fitted", xlab = "Fitted", ylab = "Average Playtime")
abline(a=0, b=1, lty=2)
```

To check the conditional mean response assumption, we analyze the
Response vs. Fitted scatterplot @fig-resp-fit. The data points are
tightly clustered in the bottom-left corner, which satisfies this
assumption. The fitted line generally follows the line $y=\hat{y}$, but
there are notable outliers along the Y-axis where the observed
`average_playtime` deviates significantly from the predicted values.
These extreme outliers suggest that the response variable may be highly
skewed or contain problematic data points.

@@

<!-- ## Create Pairwise Scatterplots for Predictors to look for multicollinearity -->

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-pairwise
#| fig-cap: "Pairwise plot between numeric predictors"

@@

<!-- ## Create Residuals vs Fitted Scatterplot to check Linearity, Uncorrelated Errors, and Constant Variance -->

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-resfit
#| fig-cap: "Residual vs fitted plot for model lm(average_playtime ~ price + positive_ratings + 
#|  negative_ratings + achievements + owners, data=data)"

@@

<!-- ## Create Normal QQ Plot to check Normality assumption -->

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-qq
#| fig-cap: "Residual vs fitted plot for model"

qqnorm(y_value)
qqline(y_value)
```

<!-- ## Box-Cox Transformation -->

```{r}
#| echo: false
#| warning: false
#| message: false

@@

<!-- ## Fitting transformed model -->

```{r}
#| echo: false
#| warning: false
#| message: false

@@

<!-- ## Checking for Multicollinearity -->

```{r}
#| include: false
#| echo: false
#| warning: false
#| message: false

vif(model_transformed)
```

To address this issue, `bc_negative_ratings` was removed from the model,
leaving the final predictors as `bc_price` (log-transformed of `price`),
`bc_positive_ratings`(log-transformed of `positive_ratings`),
`bc_achievements`(log-transformed of `achievements`) and `owners`. The
refined model was re-evaluated with summary statistics and residual
plots. The summary of coefficients as shown in @tbl-summary will be
discussed in Section @sec-conclusion. The residual diagnostics
@fig-reduced show significant improvement, with a more random
distribution around zero in the Residuals vs. Fitted plot and a QQ plot
with better alignment to the reference line.

<!-- ## Generatingn Residual Plots -->

```{r, fig.width=8, fig.height=6}
#| echo: false
#| warning: false
#| message: false
#| label: fig-reduced
#| fig-cap: "Diagnostic plots for lm(bc_average_playtime ~ bc_price + bc_positive_ratings + bc_achievements + owners, data=data)"

@@

# plot(x = x_value, y = y_value, main = "Residual vs. Fitted", xlab = "Fitted", ylab = "Residuals")
```

To further simplify the model, automated selection procedures using AIC
in both forward and backward directions were conducted. The procedure
resulted in the same final model as described above, confirming its
robustness.

<!-- ## AIC Model Selection -->

```{r}
#| include: false
#| echo: false
#| warning: false
#| message: false

step_model <- stepAIC(reduced, direction = "both")
summary(step_model)
```

<!-- ## ANOVA Comparison between full model and reduced model -->

@@

In conclusion, the model demonstrates that the average playtime of a
game is positively influenced by several factors, including the game's
price, the number of achievements, the number of positive reviews, and
the estimated number of owners. Notably, the number of owners has the
most substantial effect, as shown in the model summary @tbl-summary. For
instance, games with an estimated ownership level of 100,000–200,000
have a coefficient of approximately 0.81; as ownership grows by a factor
of 10, the effect increases to \~1.70, and for another 10-fold increase,
the effect grows to \~3.04. This indicates that games with a larger
player base tend to have higher average playtime, likely due to
increased community engagement and multiplayer opportunities.

@@

This study faces limitations primarily due to assumption violations and
data issues. The Residuals vs. Fitted plot revealed a fanning pattern,
indicating heteroscedasticity, while the Normal QQ Plot showed
significant non-normality. Outliers in the response variable
`average_playtime` suggest skewed data, despite transformations like
Box-Cox improving stability. Multicollinearity between
`bc_positive_ratings` and `bc_negative_ratings` required removing the
latter, though some interdependence remains.

The dataset, limited to Steam games, introduces platform-specific
biases, restricting generalizability. Observational data prevents causal
inference, and omitted variables, such as marketing strategies, likely
impact playtime. Future studies should refine transformations, address
multicollinearity more rigorously, and expand variables and platforms to
enhance the robustness and applicability of the findings.

# Ethics discussion

The Steam Games dataset [@davis2023steam] used in this research contains
metadata on over 97,000 games published on the Steam platform. The
dataset is sourced transparently from Steam store pages and the Steam
API, hosted on Kaggle, and widely used in existing literature, ensuring
its credibility.

In our analysis, we chose automated selection methods due to the
dataset's size and complexity. Manual selection methods were not used as
they would have been impractical and prone to human error, reducing
consistency and reproducibility. While both methods are ethically
equivalent when executed carefully, automated methods better align with
the ethical principles outlined in the second ethics module by
minimizing foreseeable risks and ensuring reliability.

Avoiding blameworthy practices requires addressing foreseeability,
avoidability, and potential errors. Automated methods allowed us to
mitigate these risks efficiently, ensuring accountability in our
approach. Manual methods, while suitable for smaller datasets, would
have introduced avoidable errors and diminished reliability, which could
be considered ethically negligent.

By prioritizing virtues such as precision and accountability, we ensured
an ethically sound methodology. Our choice was guided by practical
considerations without compromising ethical standards, fostering robust
and reproducible results.

@@

# Appendix {.unnumbered}

# Model details {#sec-model-detail}

## Residuals vs each Quantitative Predictor scatterplot to check Linearity, Uncorrelated Errors, and Constant Variance

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-respre
#| fig-cap: "Residuals vs each Quantitative Predictor"

@@

## BoxPlot to check assumptions for Categorical Predictors

See figure @fig-boc-cat

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boc-cat
#| fig-cap: "Boxplot residual vs. Owner levels suggest no linear assumption violation regarding the categorical predictors"

@@

boxplot(y_value ~ data$owners , main="Residual vs Owners",
xlab="Owners", ylab="Residuals")
```

## Final model results

<!-- ## Fitting Reduced Model -->

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summary

@@

Checking for problematic points-leverage cutoff: \~0.006, cook's
distance cutoff: \~0.961, dffits cutoff: \~0.105, dfbeta cutoff:
\~0.025. Number of influential points by each case-leverage points: 216,
regression outliers: 7, Cook's distance: 0, DFFITS: 211, by Beta from 0
to 7: 431, 476, 377, 394, 464, 321, 49, 0.
